A tutorial on Generalised Additive Mixed Effects Models for bilingualism research

### 2023-12-14 manuscript/manuscript.qmd
manuscript draft 1.0

Quantitative analysis of linguistic data has seen a massive shift from statistical significance tests and analysis of variance to linear and generalised linear modelling, including models with so-called random effects (also known as hierarchical, mixed-effects, multi-level or nested models) [@winter2013; @winter2020]. Soon after linear models started being more commonly used, a need for models that could handle non-linear effects and were as flexible as linear models led to the adoption of Generalised Additive Linear (Mixed) Models, or GAMMs.

GAMMs are now commonly employed in phonetic research (given the naturally dynamic nature of speech data) and this is reflected by the availability of several tutorials which focus on phonetic data [@soskuthy2017; @soskuthy2021; @wieling2018; @tamminga2016]. This is why we hope that this tutorial will make GAMMs accessible to researchers from other fields within linguistics. In particular, this tutorial is thought for researchers in bilingualism and multilingualism who wish to be able to start using GAMMs for non-linear data, which is very common in developmental and learning phenomena.

@@

Generalised Additive Models (GAMs) and their mixed-effects version (GAMMs) are a generalisation of generalised linear models that can accommodate fitting non-linear models [@hastie1986; @wood2006]. The main component that is specific to GAMMs are smoothers or smooth terms. These are model terms that can fit non-linear effects. An advantage of using smooth terms is that they are constrained by a penalty measure---part of the fitting procedure and does not depend on user input---which prevents over-fitting the data. This approach is different from, for example, growth curve analyses which are based on user-specified polynomial degrees.

Readers interested in a more detailed conceptual discussion of GAMMs are referred to @soskuthy2017. The rest of this tutorial will focus on teaching the basics of GAMMs usage in R. While completing this tutorial will allow you to start-jump into fitting basic GAMMs, we invite interested readers to peruse the tutorials mentioned above to avoid unnecessary repetition of content. Finally, we opted not to treat statistical significance testing in GAMMs, due to its cumbersome nature (significance testing is treated in details in the aforementioned tutorials).

@@

@@

You can find the tutorial code and data here: XXX.

@@

```{r}
#| label: dat1
# Print first 6 rows of the dataframe
head(dat1)
```

@@

```{r}
#| label: fig-dat1
#| echo: false

@@

```{r}
#| label: fig-gam-1

@@

As mentioned previously, the data `dat1` contains accuracy and proficiency scores from 200 subjects, taken at 10 time points. The data was simulated so that proficiency increased over time (at different degrees for different speakers), as is often the case in language acquisition processes. An interesting question to consider is whether learning scores improve with time independently of proficiency, or if proficiency alone is causing learning scores to improve. We can approach this question using principles of causal inference. To this end, we develop directed acyclic graphs (DAGs) to describe the causal relationships involved. A full treatment of causal inference is beyond the scope of this paper. Interested readers are referred to XXX.

@@

...

Now we plot the model predictions.

@@

gam_2_preds <- predict_gam(
  gam_2, length_out = 25,
  series = "proficiency",
  exclude_terms = c("s(subj)", "s(subj,proficiency)")
)

@@

gam_2_preds_t <- predict_gam(
  gam_2, length_out = 25,
  series = "time",
  exclude_terms = c("s(subj)", "s(subj,proficiency)")
)

@@

In the following example, we consider the production of Spanish vowels in the aforementioned groups of bilingual speakers. Interestingly, to our knowledge, all of the current research investigating the cross-linguistic influence of vowel reduction processes in learners of Spanish have utilized modelling strategies that rely upon formant measurements at a single time point, typically the mid-point of the vowel, rather than scrutinizing the entire trajectory of the spectral envelope. In our view, the dynamic nature of vowel formants poses an interesting use-case for modelling using non-linear methods, such as GAMs. To this end, we have simulated phonetic data from three aforementioned groups of bilinguals: simultaneous English/Spanish bilinguals, beginner late English/Spanish bilinguals and advanced late English/Spanish bilinguals. The data contains the Euclidean Distance (EuD) of three Spanish vowels /a, i, u/ from the vowel space centroid, taken from word-final non-stressed vowels, along nine equidistant time points. The measure is a good proxy for vowel reduction phenomena, which is to be expected in late bilinguals.

@@

dat2
```

@@

```{r}
#| label: fig-dat2
#| echo: false

@@

Furthermore, `predict_gam()` returns an object of class `tidygam` which can be plotted with `plot()` (similarly to how `ggpredict()` from the ggeffects package works, XXX). However, we might want to do some processing of the predictions before plotting: in particular, we might want to split the `group_vowel` column into the two original variables, `group` and `vowel`. We can achieve this straight from the function. And we might also want to extract more than 10 time points, to get a smoother curve.

@@

# References

>>>>

In recent years quantitative analysis of linguistic data has benefited greatly from the now commonplace use of so-called random effects (also known as hierarchical, mixed-effects, multi-level or nested models, @winter2013; @winter2020). While this shift represents a clear improvement in statistical analyses, particularly in areas of linguistics in which nested structure is the norm (e.g., trial repetitions nested within participants), an over-reliance on standard linear modelling prevails, particularly in the cases of dynamic phenomena that may not constitute a linear relationship, e.g., vowel trajectories, pitch contours, acquisition processes, etc.

Generalised Additive (Mixed) Models (GAMMs) are now commonly employed in phonetic research (given the naturally dynamic nature of speech data) and this is reflected by the availability of several tutorials which focus on phonetic data [@soskuthy2017; @soskuthy2021; @wieling2018; @tamminga2016]. This is why we hope that this tutorial will make GAMMs accessible to researchers from other fields within linguistics. In particular, this tutorial is thought for researchers in bilingualism and multilingualism who wish to be able to start using GAMMs for non-linear data, which is very common in developmental and learning phenomena.

@@

Generalised Additive Models (GAMs) and their mixed-effects version (GAMMs) represent a variation of the generalised linear model that can accommodate modelling non-linear relationships [@hastie1986; @wood2006]. The main component that is specific to GAMMs are smoothers or smooth terms. These are model terms that can fit non-linear effects. An advantage of using smooth terms is that they are constrained by a penalty measure---part of the fitting procedure and does not depend on user input---which prevents over-fitting the data. This approach is different from, for example, growth curve analyses which are based on user-specified polynomial degrees.

Readers interested in a more detailed conceptual discussion of GAMMs are referred to @soskuthy2017. The rest of this tutorial focuses on applying GAMMs in bilingualism research using the statistical programming language R [@rct2023].

<!-- While completing this tutorial will allow you to jump-start into fitting basic GAMMs, we invite interested readers to peruse the tutorials mentioned above to avoid unnecessary repetition of content. Finally, we opted not to treat statistical significance testing in GAMMs, due to its cumbersome nature (significance testing is treated in details in the aforementioned tutorials). -->

@@

In what follows, we detail the packages needed to complete the tutorial, followed by two specific case studies designed to illustrate how GAMMs can be applied to model common phenomena in second language acquisition and bilingualism research.

@@

You can find the tutorial code and data here: <https://github.com/stefanocoretta/gamm_biling>.

@@

```{r}
#| label: dat1
dat1 <- readRDS("data/dat1.rds")

# Print first 6 rows of the dataframe
head(dat1)
```

@@

```{r}
#| label: fig-dat1
#| echo: false
#| fig-cap: "Scatter plot of proficiency and learning scores illustrating a U-shaped learning curve."

@@

```{r}
#| label: fig-gam-1
#| fig-cap: "Predicted accuracy score depending on proficiency, from a GAM."

@@

As mentioned previously, the data `dat1` contains accuracy and proficiency scores from 200 subjects, taken at 10 time points. The data was simulated so that proficiency increased over time (at different degrees for different participants), as is often the case in language acquisition processes. An interesting question to consider is whether learning scores improve with time independently of proficiency, or if proficiency alone is causing learning scores to improve. We can approach this question using principles of causal inference. To this end, we develop directed acyclic graphs (DAGs) to describe the causal relationships involved. A full treatment of causal inference is beyond the scope of this paper. Interested readers are referred to @mcelreath2019.

@@

Now we plot the model predictions of accuracy score by proficiency and time point separately.

@@

gam_2_preds <- predict_gam(
  gam_2, length_out = 25,
  series = "proficiency",
  exclude_terms = c("s(subj)", "s(subj,proficiency)")
)
```

```{r}
#| label: fig-gam-2-preds
#| fig-cap: "Predicted accuracy score as a function of proficiency, from a GAM."

@@

gam_2_preds_t <- predict_gam(
  gam_2, length_out = 25,
  series = "time",
  exclude_terms = c("s(subj)", "s(subj,proficiency)")
)
```

```{r}
#| label: fig-gam-2-preds-t
#| fig-cap: "Predicted accuracy score as a function of proficiency, from a GAM."

@@

In the following example, we consider the production of Spanish vowels in the aforementioned groups of bilingual participants. Interestingly, to our knowledge, all of the current research investigating the cross-linguistic influence of vowel reduction processes in learners of Spanish have utilized modelling strategies that rely upon formant measurements at a single time point, typically the mid-point of the vowel, rather than scrutinizing the entire trajectory of the spectral envelope. In our view, the dynamic nature of vowel formants poses an interesting use-case for modelling using non-linear methods, such as GAMs. To this end, we have simulated phonetic data from three aforementioned groups of bilinguals: simultaneous English/Spanish bilinguals, beginner late English/Spanish bilinguals and advanced late English/Spanish bilinguals. The data contains the Euclidean Distance (EuD) of three Spanish vowels /a, i, u/ from the vowel space centroid, taken from word-final non-stressed vowels, along nine equidistant time points. The measure is a good proxy for vowel reduction phenomena, which is to be expected in late bilinguals.

@@

dat2 <- readRDS("data/dat2.rds")

head(dat2)
```

@@

```{r}
#| label: fig-dat2
#| echo: false
#| fig-cap: "Euclidean Distance through 9 time points taken from within vowels."

@@

Furthermore, `predict_gam()` returns an object of class `tidygam` which can be plotted with `plot()` (similarly to how `ggpredict()` from the ggeffects package works, @luedecke2018). However, we might want to do some processing of the predictions before plotting: in particular, we might want to split the `group_vowel` column into the two original variables, `group` and `vowel`. We can achieve this straight from the function. And we might also want to extract more than 10 time points, to get a smoother curve.

@@

## References
